Verification of Stencil Computations through Spatial Specifications

### 2016-07-05 paper.tex
removing spectop references

laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+ &
  (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
\end{minted}
(a) Excerpt from Fortran code for a Navier-Stoke fluid simulator,
showing highly-detailed stencil computations. \\

%\begin{SpecVerbatim}[xleftmargin=0.1cm]
%20-24  u: centered depth=1 dim=1
%6-30  u: centered depth=1 dim=2
%      v: forward depth=1 dim=1, backward depth=1 dim=1
%32-33  u: centered depth=1 dim=1,2
%\end{SpecVerbatim}

\begin{minted}{fortran}
!= stencil centered(depth=1,dim=1)*reflexive(dim=2) + centered(depth=1,dim=2)*reflexive(dim=1) :: u

!= stencil forward(depth=1,dim=1) * backward(depth=1,dim=2) :: v
\end{minted}
(b) Inferred stencil specification from CamFort

\begin{center}
\begin{tikzpicture}
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}
(c) Pictorial representation of the two stencil specifications.
The horizontal dimensions is \text{dim=1} and the vertical is \texttt{dim=2}:
\caption{Fragment of Navier-Stokes fluid simulator and its specification}
\label{ref:navier-stokes-fragment}
\end{figure}

@@

The last six equations (from (\textsc{\term{+}idem}) inclusive), explain
the algebraic behaviour of \term{+} and \term{*}.  Together, we see
that \term{+} is an associative, commutative, idempotent binary
operator that distributes with \term{*}, which is associative and
commutative. This distribution of \term{*} over \term{+} is key as it
is used internally to give a normal form for stencil specifications
akin to Disjunctive Normal Form (DNF) (where \term{+} is taken as
``disjunction'' and \term{*} as the ``conjunction''). We revisit this
normalisation in Section~\ref{}.

@@

We demonstrate the main steps of the inference procedure over the
following example program which computes the mean value
of a five-point stencil at each index of the input array:
%%
\begin{minted}{fortran}
do i = 1, (n-1)
   do j = 1, (m-1)
      x       = a(i-1, j) + a(i+1, j)
      y       = a(i, j-1) + a(i, j+1)
      b(i, j) = (a(i, j) + x + y) / 5.0
   end do
end do
\end{minted}
%%
\paragraph{Step 1: Standard control and data-flow analyses}
The inference relies on some standard program analyses, computed
before the main inference procedure:
%
\begin{enumerate}
\item basic blocks (CFG);
\item induction variables per basic block;
\item (interprocedural) data-flow analysis, providing a \emph{flows to}
  graph (as shorthand, the function
  $\mathsf{flowTo}$ is used, implicitly parameterised by this graph,
  mapping an expression to the set of all expressions
  with forwards data-flow to this expression, based on the transitive
  closure of the flows graph);
\item type information per variable, where we use the predicate
\arrayTy{} to classify variables of array type.
\end{enumerate}
%

\paragraph{Step 2: Data-access analysis}

@@

\paragraph{Step 3: Coalesce contiguous indices into regions}

@@

\paragraph{Step 4: Convert spans into region AST}

Given the set $\textsf{regions}(M)$ of spans covering the indexing
space, the next step converts these into abstract syntax trees
for the specification language. This conversion is defined in terms
of an algebra on \textit{specTop} syntax which mirrors the shape
of the syntax with additive and multiplicative binary operators
$\oplus$ and $\odot$ and a constructor for \textit{specTop}:
%%
\begin{align*}
\oplus & : \textit{specTop} \rightarrow \textit{specTop} \rightarrow
  \textit{specTop} \\
\odot & : \textit{specTop} \rightarrow \textit{specTop} \rightarrow
        \textit{specTop} \\
\mathsf{toSpec} & : (\mathbb{Z} \cup \{\infty\}) \rightarrow (\mathbb{Z} \cup
  \{\infty\})  \rightarrow \textit{specTop}
\end{align*}
%%
The operation $\mathsf{toSpec}$ takes a pair of lower and upper bounds
values and constructs a specification which contains a one-dimensional
region. This step of the inference then proceeds by converting each
span into a $n$-dimensional specification by converting each
pair of bounds in a span and multiplying by $\otimes$
then summing the specifications resulting specification by $\oplus$.
That is: 
\begin{equation*}
\textit{spec}(M) = 
\sum\limits_{\vect{x} \in \textsf{regions}(M)}
\prod\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec}(\vect{x}^L_i, \vect{x}^U_i)
\end{equation*}

For each span, and each dimension within it, each pair of lower and
upper bounds is mapped into a specification. That is,
for all $\vect{x}$ of size $n$, and for all $i \in \{1,\ldots,n\}$
then $\mathsf{toSpec} (\vect{x}^L_i, \vect{x}^U_i)$ maps each
pair of lower and upper bounds in each dimension to a piece of syntax 
of \textit{specTop} syntax, defined:
%%
%\newcommand{\llower}{\vect{x}^L_i}
%\newcommand{\lupper}{\vect{x}^U_i}
\newcommand{\llower}{l}
\newcommand{\lupper}{u}
\begin{align*}
&\mathsf{toSpec} \; (\llower, \lupper) = \\
&\begin{cases}
%% FWD
 \stenFwdS{\lupper}{i}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.3em]
 \stenFwdSR{\lupper}{i}{\irreflS}  & \llower = 1 \, \wedge \, \lupper > 0 \\[0.3em]
%% BWD
 \stenBwdS{|\llower|}{i}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.3em]
\stenBwdSR{|\llower|}{i}{\irreflS} & \llower < 0 \, \wedge \, \lupper = -1 \\[0.3em]
%% CEN
 \stenCenS{\lupper}{i}  & \llower < 0 \, \wedge \, \lupper > 0
 \, \wedge \, |\llower| = \lupper \\[0.3em]
 \stenBwdS{|\llower|}{i} & \llower < 0 \, \wedge \, \lupper > 0  \, \wedge \, |\llower| \neq \lupper \\
\quad \texttt{+} \, \stenFwdS{\lupper}{i}  & \qquad\quad \\[0.3em]
% ABSOLUTE REP
 \epsilon & \llower = \infty \, \vee \, \lupper
 = \infty \\[0.3em]
% BOUND
 \texttt{atMost}, \stenFwdS{\lupper}{i} & \llower > 1 \\[0.3em]
 \texttt{atMost}, \stenBwdS{|\llower|}{i} \quad & \llower < (-1) \\[0.3em]
  \end{cases}
\end{align*}
%%
The first two cases map to \texttt{forward} stencils since the lower
bound is $0$ or $1$, with the depth determined by the upper bound of
the span. If the lower bound is $1$, then the
$0$-offset point is missing so the region is given an
\texttt{irreflexive} attribute. The second two cases are similar for
\texttt{backward} but now the depth of the region is determined by the
absolute of the lower bound. In the fifth-case, where the lower bound
is negative and the upper bound is positive and they have the same
magnitude, then a \texttt{centered} region is constructed with
the depth as the common magnitude. If the magnitudes are not the same
(sixth case) then separated \texttt{forward} and \texttt{backward}
regions are summed, with their corresponding depths.

In the case where either bound is $\infty$, representing a non
neighbourhood index, then an empty specification is given
 with no regions, indicating that there is not even reflexive
0-offset access in this dimension.

Finally in the last two cases, if the lower bound is greater than 1
(which implies the upper bound is greater than 1 too) then
this represents a region that is not at the origin (or next to) the
origin and thus cannot be represented in the specification language.
Thus, an upper bound is instead returned showing the maximum extent
of the region. The last case is the dual, when the lower bound is less
than $-1$, giving an upper bound on a \texttt{backward} specification.

\begin{lemma}
$\mathsf{toSpec}_i$ is total (\ie{}, defined for all spans) given the precondition that the lower
bound of the input span is less than or equal to its upper bound.
\end{lemma}

@@

>>>>

laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+ &
  (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
\end{minted}
(a) Excerpt from Fortran code for a Navier-Stoke fluid simulator,
showing highly-detailed stencil computations. 
\vspace{0.5em}
\hrule
%
\begin{minted}{fortran}
!= stencil centered(depth=1,dim=1)*reflexive(dim=2) + centered(depth=1,dim=2)*reflexive(dim=1) :: u

!= stencil forward(depth=1,dim=1) * backward(depth=1,dim=2) :: v
\end{minted}
(b) Inferred stencil specification from CamFort.
\vspace{0.5em}
\hrule
%
\begin{center}
\begin{tikzpicture}
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}
\vspace{-0.3em}
(c) Pictorial representation of the two stencil specifications.
The horizontal dimension is \texttt{dim=1} and the vertical is \texttt{dim=2}.
\caption{Fragment of a Navier-Stokes fluid simulator and its
  specification in our language}
\label{ref:navier-stokes-fragment}
\end{figure}

@@

The last six equations (from (\textsc{\term{+}idem}) inclusive), explain
the algebraic behaviour of \term{+} and \term{*}.  Together, we see
that \term{+} is an associative, commutative, idempotent binary
operator that distributes with \term{*}, which is associative and
commutative. This distribution of \term{*} over \term{+} is key as it
is used internally to give a normal form for stencil specifications
akin to Disjunctive Normal Form (DNF) (where \term{+} is taken as
``disjunction'' and \term{*} as the ``conjunction''). We revisit this
normalisation in Section~\ref{sec:inf-step4}.

@@

We demonstrate the main steps of the inference procedure over the
following example program which computes the mean value
of a five-point stencil at each index of the input array:
%%
\begin{minted}{fortran}
do i = 1, (n-1)
   do j = 1, (m-1)
      x       = a(i-1, j) + a(i+1, j)
      y       = a(i, j-1) + a(i, j+1)
      b(i, j) = (a(i, j) + x + y) / 5.0
   end do
end do
\end{minted}
%%
\subsubsection{Step 1: Standard control and data-flow analyses}
\label{sec:inf-step1}

The inference relies on some standard program analyses, computed
before the main inference procedure:
%
\begin{enumerate}
\item basic blocks (CFG);
\item induction variables per basic block;
\item (interprocedural) data-flow analysis, providing a \emph{flows to}
  graph (as shorthand, the function
  $\mathsf{flowTo}$ is used, implicitly parameterised by this graph,
  mapping an expression to the set of all expressions
  with forwards data-flow to this expression, based on the transitive
  closure of the flows graph);
\item type information per variable, where we use the predicate
\arrayTy{} to classify variables of array type.
\end{enumerate}
%

\subsubsection{Step 2: Data-access analysis}
\label{sec:inf-step2}

@@

\subsubsection{Step 3: Coalesce contiguous indices into regions}
\label{sec:inf-step3}

@@

\subsubsection{Step 4: Convert spans into specifications}
\label{sec:inf-step4}

\newcommand{\bplus}{\operatornamewithlimits{\term{+}}}
\newcommand{\tySum}[1]{#1^{\term{+}}}
\newcommand{\tyProd}[1]{#1^{\term{*}}}
\newcommand{\specDNF}{\textbf{spec}}

Given the set $\textsf{regions}(M)$ of spans covering the indexing
space, the next step converts these into abstract syntax trees
for the specification language. This conversion is defined in terms
of an algebra on \textit{spec} syntax in disjunctive-normal form
(which we denote \specDNF{}). The algebra mirrors the shape
of the region syntax with additive and multiplicative binary operators
$\oplus$ and $\odot$ and a constructor for \specDNF{}:
%%
\begin{align*}
\oplus & : \specDNF \rightarrow \specDNF \rightarrow
  \specDNF \\
\odot & : \specDNF \rightarrow \specDNF \rightarrow
        \specDNF \\
\mathsf{toSpec} & : (\mathbb{Z} \cup \{\infty\}) \rightarrow (\mathbb{Z} \cup
  \{\infty\})  \rightarrow \specDNF
\end{align*}
%%
The $\mathsf{toSpec}$ operation takes a pair of lower and upper bounds
values (one-dimensional) and constructs a specification (in the
\textit{spec} syntax) which contains a one-dimensional
region. An entire span is convereted then into a $n$-dimensional
specification by applying $\mathsf{toSpec}$ to each
pair of bounds in a span, per dimension, and multiplying these
specifications by $\odot$. The resulting specifications for
each span are then summed by by $\oplus$, that is:
\begin{equation*}
\textit{spec}(M) = 
\sum\limits_{\vect{x} \in \textsf{regions}(M)}
\prod\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec}(\vect{x}^L_i, \vect{x}^U_i)
\end{equation*}
(where $\oplus$ is the additive and $\odot$ is the multiplicative used
within $\Sigma$ and $\Pi$ respectively here).

\begin{definition}[DNF specifications]
Let $r_1, ..., r_n$ range over elements of the \textit{region}
syntax (see Figure~\ref{fig:syntax}, p.~\pageref{fig:syntax})
The domain of \specDNF{} specifications
in disjunctive normal form ranges over terms of the form:
%%
(r_1 \term{*} \ldots \term{*} r_n)\term{+} \ldots 
\term{+} (r_1 \term{*} \ldots \term{*} r_n)
%%
\end{definition}

\begin{definition} The $\mathsf{toSpec}$ constructor is defined:
%%
%\newcommand{\llower}{\vect{x}^L_i}
%\newcommand{\lupper}{\vect{x}^U_i}
\newcommand{\llower}{l}
\newcommand{\lupper}{u}
\begin{align*}
&\mathsf{toSpec} \; \llower \; \lupper = \\
&\begin{cases}
%% FWD
 \stenFwdS{\lupper}{i}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.2em]
 \stenFwdSR{\lupper}{i}{\irreflS}  & \llower = 1 \, \wedge \, \lupper > 0 \\[0.2em]
%% BWD
 \stenBwdS{|\llower|}{i}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.2em]
\stenBwdSR{|\llower|}{i}{\irreflS} & \llower < 0 \, \wedge \, \lupper = -1 \\[0.2em]
%% CEN
 \stenCenS{\lupper}{i}  & \llower < 0 \, \wedge \, \lupper > 0
 \, \wedge \, |\llower| = \lupper \\[0.2em]
 \stenBwdS{|\llower|}{i} & \llower < 0 \, \wedge \, \lupper > 0  \, \wedge \, |\llower| \neq \lupper \\
\quad \texttt{+} \, \stenFwdS{\lupper}{i}  & \qquad\quad \\[0.2em]
% BOUND
 \texttt{atMost}, \stenFwdS{\lupper}{i} & \llower > 1 \\[0.2em]
 \texttt{atMost}, \stenBwdS{|\llower|}{i} \quad & \llower < (-1)
 \\[0.2em]
% ABSOLUTE REP
 \epsilon & \llower = \infty \, \vee \, \lupper = \infty
  \end{cases}
\end{align*}
%%
The first two cases map to \texttt{forward} stencils since the lower
bound is $0$ or $1$, with the depth determined by the upper bound of
the span. If the lower bound is $1$, then the
$0$-offset point is missing so the region is given an
\texttt{irreflexive} attribute. The second two cases are similar for
\texttt{backward} but now the depth of the region is determined by the
absolute of the lower bound. In the fifth-case, where the lower bound
is negative and the upper bound is positive and they have the same
magnitude, then a \texttt{centered} region is constructed with
the depth as the common magnitude. If the magnitudes are not the same
(sixth case) then separated \texttt{forward} and \texttt{backward}
regions are summed, with their corresponding depths.

If the lower bound is greater than 1
(which implies the upper bound is greater than 1 too) then
this represents a region that is not at the origin (or next to) the
origin and thus cannot be represented in the specification language.
Thus, an upper bound is instead returned showing the maximum extent
of the region. The penultimate case is the dual, when the lower bound is less
than $-1$, giving an upper bound on a \texttt{backward} specification.

Finally, in the case where either bound is $\infty$, representing a non
neighbourhood index, then an empty specification is given
 with no regions, indicating that there is not even reflexive
0-offset access in this dimension.
\end{definition}

\begin{lemma}
$\mathsf{toSpec}$ is total (\ie{}, defined for all inputs)
 given the precondition that the lower
bound of the input span is less than or equal to its upper bound.
\end{lemma}

%\newcommand{\mkBig}[2]{\overset{\LARGE{\term{#1}}}{#2}}
\begin{definition} The $\odot$ operation is defined:
\begin{align*}
\tySum{r_1} \odot \tySum{r_2} = 
 \bplus_{(\tyProd{s_1}, \tyProd{s_2}) \in (\tySum{r_1} \times
   \tySum{r_2})} (\tyProd{s_1} \; \term{*} \; \tyProd{s_2})
%\begin{cases}%
%\end{cases}
\end{align*}
\end{definition}

@@

\textbf{All reflexive} The vast majority (X\%) of stencils were
reflexive in every dimension of the array. These most often correspond
to applying point-wise transformations to data such as scaling
e.g. \begin{minted}{fortran}
ZZ1(I,K) = .000041868*ZZ1(I,K)
\end{minted}
